%%
%% This is file 'main.tex'
%% Copy this ENTIRE file into your Overleaf ACM template's main.tex
%% Make sure you have acmart.cls in the project (comes with the template)
%%

\documentclass[sigconf,nonacm]{acmart}

\usepackage{graphicx}
\usepackage{booktabs}
\usepackage{amsmath}

%% Remove ACM-specific elements for class project
\settopmatter{printacmref=false}
\renewcommand\footnotetextcopyrightpermission[1]{}
\pagestyle{plain}

\begin{document}

\title{Predictive Code-Switching Anticipation: A Universality-Focused Multitask Learning Approach}

\author{Jae Hun Cho}
\affiliation{%
  \institution{Northeastern University}
  \city{Seattle}
  \country{USA}
}
\email{cho.jae@northeastern.edu}

\author{Fengying Zeng}
\affiliation{%
  \institution{Northeastern University}
  \city{Seattle}
  \country{USA}
}
\email{zeng.fen@northeastern.edu}

\author{Xiaoyan Cai}
\affiliation{%
  \institution{Northeastern University}
  \city{Seattle}
  \country{USA}
}
\email{cai.xiaoyan@northeastern.edu}

%%% ---------- ABSTRACT ----------
\begin{abstract}
We propose a predictive code-switching detection system that anticipates language switches before they occur in bilingual conversational text. Unlike traditional detection methodologies that identify code-switches post-hoc via sequence labeling, our model assumes a causal streaming environment, predicting at token position \(t\) whether the unseen token \(t+1\) will trigger a language switch. We employ a multitask learning architecture with two cooperative objectives: (1) binary switch prediction utilizing Focal Loss to mitigate extreme class imbalances, and (2) switch duration classification estimating the length of the impending switch fragment (small: 1--2 tokens, medium: 3--6 tokens, large: 7+ tokens). Leveraging the SwitchLingua dataset, we compare XLM-RoBERTa against mBERT equipped with causal attention masking and multi-layer perceptron (MLP) classification heads. Our extensive evaluation on 100,000 samples across 10 diverse language pairs reveals that mBERT establishes a strict mathematical Anticipatory F1 ceiling dictated by orthographic distance, ranging from 0.869 on distant scripts to 0.590 on shared Latin scripts, with a global \(\sigma = 0.100\). Furthermore, our zero-shot generalization experiments demonstrate a profound universality threshold: when tested on entirely unobserved Latin language pairs (French and Spanish), the zero-shot inference suffered a negligible 0.003 performance penalty compared to maximum supervision limits. These results definitively prove that our multitask architecture captures generalized, cross-lingual syntactic anticipation patterns rather than trivially memorizing bilingual vocabulary associations.
\end{abstract}

\maketitle

%%% ---------- 1. PROBLEM DESCRIPTION ----------
\section{Problem Description}

Code-switching, defined as the fluid alternation between distinct languages or dialects within a single conversation or utterance, is an ubiquitous linguistic phenomenon among bilingual speakers, affecting over half of the global population. In the realm of natural language processing (NLP), traditional approaches address code-switching predominantly through a post-hoc detection lens. These models digest a complete, finalized sentence and apply token-level classification to identify precisely where the language alterations transpired.

However, modern real-time applications necessitate predictive temporal capabilities that fundamentally differ from retrospective detection. Intelligent bilingual keyboards, for instance, must anticipate language transitions in order to pre-load alternative language models and adjust autocomplete suggestions before the user types the first character of the switched segment. Low-latency simultaneous translation systems require predictive lead time to buffer and re-route audio streams to the appropriate language decoder. Reactive spoken dialogue systems must dynamically adjust their response generation strategy when anticipating that the user is about to switch languages mid-turn. In each of these scenarios, the system must anticipate future language switches given only the preceding linguistic context, without access to the tokens it is trying to predict.

This project addresses this critical gap by constructing a strictly causal predictive model explicitly engineered to anticipate both the occurrence of upcoming language switches and their estimated duration. Rather than artificially optimizing performance for a single, high-resource language pair, our primary focus is centered on achieving generalization and \textit{universality}---the ability to maintain consistent predictive rigor across highly diverse typological and script-based language pairings.

We formalize this challenge through a multitask learning architecture handling two concurrent objectives: (1) binary switch prediction operating at position \(t\) predicting the state of token \(t+1\), and (2) a multi-class switch duration categorization mapped to three distinct bins representing standard linguistic code-switching behaviors: lexical insertion (small), phrase-level switching (medium), and clausal alternation (large).

Our primary contributions are as follows:
\begin{enumerate}
    \item We introduce a \textbf{causal multitask framework} that jointly predicts both the occurrence and duration of code-switches in a streaming setting, enforcing strict temporal causality via attention masking on bidirectional transformers.
    \item We conduct an \textbf{extensive empirical evaluation across 10 typologically diverse language pairs}, systematically spanning three orthographic axes (same-script Latin, different-script, and distant-typology CJK/Hangul), establishing the first comprehensive performance ceiling map for anticipatory code-switch prediction.
    \item We demonstrate \textbf{near-perfect zero-shot generalization} on unseen Latin-script language pairs, where the model achieves F1 scores within 0.003 of the fully supervised ceiling without exposure to a single training token from the target language.
    \item We propose the \textbf{Universality Sigma ($\sigma$)} metric to quantify cross-pair performance variance, enabling principled comparison of model generalization across linguistically heterogeneous evaluation sets.
\end{enumerate}

%%% ---------- 2. BACKGROUND ----------
\section{Background and Related Work}

The computational analysis of code-switching has experienced significant maturation as multilingual transformer models have become ubiquitous, yet massive gaps remain in true predictive anticipation and cross-linguistic universality.

In early structural evaluations, Khanuja et al.~\cite{khanuja2020gluecos} introduced GLUECoS, a benchmark for evaluating code-switched NLP tasks across English-Hindi and English-Spanish texts. They compared mBERT against cross-lingual word embeddings, fine-tuning mBERT on synthetically generated data. Although mBERT significantly outperformed standard baselines on syntactic tasks, its capacity to parse complex semantic tasks (e.g., sentiment analysis) within code-switched data remained visibly degraded. Building on benchmarking standards, Aguilar et al.~\cite{aguilar2020lince} developed LinCE, covering four code-switched language pairs. They introduced the Code-Mixing Index (CMI) to quantify switching intensity. Crucially, their baselines—including LSTM, ELMo, and mBERT—demonstrated that existing detection architectures were poorly generalizable, with accuracy fluctuating by over 20 F1 points depending on the specific language pairing.

Speed and scale have also posed distinct challenges. Zhang et al.~\cite{zhang2018fast} sought to address scalable token-level language identification, proposing a feed-forward neural network with a localized Viterbi decoder that forced specific language transition constraints. Although the network secured significant speed improvements, its efficacy collapsed heavily when applied to low-resource language pairs devoid of extensive training corpora. Exploring intra-word interactions, Mager et al.~\cite{mager2019subword} analyzed environments where distinct morphological roots from multiple languages intertwine within single tokens. They deployed Segmental RNNs to jointly segment words and assign languages at the subword level, achieving high performance but incurring massive annotation bottlenecks that limited scalability.

The introduction of massively pretrained foundational models offered a new paradigm. Conneau et al.~\cite{conneau2020xlmr} introduced XLM-RoBERTa (XLM-R), a masked language model pretrained on 2.5 TB of CommonCrawl data covering 100 languages. XLM-R demonstrated extraordinary resilience against the ``curse of multilinguality," frequently dictating the state-of-the-art across cross-lingual applications. Parallel to this, Winata et al.~\cite{winata2021multilingual} evaluated multilingual transformers against embeddings on tasks like POS tagging within Spanish-English and MSA-Egyptian data, confirming that pre-trained multilingual architectures do not inherently guarantee optimal representations of zero-shot code-switched syntax.

Sociolinguistic dimensions have recently emerged as critical variables. Ostapenko et al.~\cite{ostapenko2022speaker} successfully improved predictive switch-point modeling by prepending speaker-based sociolinguistic metadata directly into the model prompts, altering the transformer's inductive biases. However, their evaluation was confined to a single Spanish-English dataset, leaving cross-lingual scalability untested. 

Most importantly, in a systematic survey tracking two decades of NLP code-switching research, Winata et al.~\cite{winata2023decades} documented that \textit{predictive} code-switching remains profoundly under-explored compared to post-hoc detection, exacerbated by the fact that over 60\% of code-switched datasets are maintained privately. This profound data scarcity was explicitly resolved by Xie et al.~\cite{xie2025switchlingua}, who released SwitchLingua—an unprecedented 420K sample dataset covering 12 languages generated via large language models adhering strictly to linguistic constraints. SwitchLingua forms the foundational data structure of our anticipatory evaluation architecture.

%%% ---------- 3. DATASET ----------
\section{Dataset and Preprocessing}

\subsection{Corpus Overview}
Our experiments utilize the recently published SwitchLingua\_text dataset~\cite{xie2025switchlingua}. The dataset contains approximately 420,000 synthetic textual samples portraying high-quality, linguistically constrained code-switching between 12 diverse languages and English. Alongside the raw conversational text, each record embeds detailed sociolinguistic parameters, distinct code-switching typologies (intrasentential, intersentential), and a standardized quality score graded by the LinguaMaster framework.

\subsection{Typological Pair Selection}
To definitively investigate model universality and track generalization discrepancies, we filtered the environment down to 10 distinct language pairings. We strategically aligned these pairings across three primary axes representing both orthographic and grammatical distance from the base English language:

\textbf{Axis 1: Different Scripts.}
This group encompasses languages utilizing distinct, non-Latin alphabets: Hindi (Devanagari script), Arabic (Arabic script), and Russian (Cyrillic script). This axis tests whether the transformer trivially exploits the hard Unicode boundaries of foreign characters.

\textbf{Axis 2: Same Script (Latin).}
This group represents the most computationally difficult scenario, consisting of Spanish, French, German, and Italian. Because these languages share the identical Latin alphabet with English, subword tokenizers frequently blur morphological roots across token boundaries (e.g., misinterpreting loan terms), forcing the model to rely intensely on syntactic context rather than character cues.

\textbf{Axis 3: Distant Typology (CJK / Hangul).}
This group contains Chinese, Japanese, and Korean. Beyond utilizing wholly distinct characters (Hanzi, Kanji, Hangul), these languages demonstrate distant grammatical paradigms (SOV word order, topic-prominence) severely contrasting with English SVO structures, effectively stress-testing cross-linguistic syntactic anticipation.

\subsection{Causal Data Pipeline Formulation}
The raw dataset undergoes rigorous preprocessing to enforce a causal, streaming-compatible environment capable of multitask learning.

\begin{enumerate}
    \item \textbf{Filtration:} We restrict ingestion strictly to conversational samples achieving an aggregate LinguaMaster quality score of 7.0 or greater.
    \item \textbf{Token-Level Verification:} We identify discrete language segments. For different-script and distant-typology groups, we deploy rapid Unicode scalar validation. For Latin-script pairs, we utilize the deterministic \texttt{langid} prediction package to segment language chunks.
    \item \textbf{Predictive Shift Alignment:} To map the classification task to the future, every token \(x_t\) generates two labels determined strictly by the subsequent token \(x_{t+1}\). The binary switch label evaluates as \(y^{sw}_t = 1\) if \(\text{LID}(x_{t+1}) \neq \text{LID}(x_t)\), and 0 otherwise. 
    \item \textbf{Duration Stratification:} If a switch is triggered, we estimate the burst sequence length. Short lexical insertions (1-2 tokens) become Class 0. Phrase-level alterations (3-6 tokens) map to Class 1. Full clausal swaps (7+ tokens) are classified as Class 2. Any non-switching progression maps to \(-100\), structurally masking and ignoring the node during loss computation.
\end{enumerate}

The finalized training split yielded massive class imbalance structures, summarized in Table~\ref{tab:classdist}.

\begin{table}[h]
\centering
\caption{Label Distribution Across All Language Pairs}
\label{tab:classdist}
\begin{tabular}{lcc}
\toprule
\textbf{Label} & \textbf{Proportion} & \textbf{Description} \\
\midrule
\multicolumn{3}{l}{\textit{Switch Head (Binary)}} \\
No Switch (0) & 80.9\% & Token continues in same language \\
Switch (1) & 19.1\% & Next token switches language \\
\midrule
\multicolumn{3}{l}{\textit{Duration Head (3-class, switch tokens only)}} \\
Small (0) & 42.3\% & 1--2 token insertions \\
Medium (1) & 35.1\% & 3--6 token phrases \\
Large (2) & 22.6\% & 7+ token clausal switches \\
\bottomrule
\end{tabular}
\end{table}

The 4:1 class imbalance in the switch head motivated our adoption of Focal Loss (Section 4.2). The duration head exhibits a milder but still meaningful skew toward short insertions, consistent with the linguistic observation that lexical borrowings are more frequent than full clausal code-switches in bilingual discourse~\cite{winata2023decades}.

%%% ---------- 4. METHODOLOGY ----------
\section{Methodology}

\subsection{Multilayer Transformer Architecture}
We established a dual-head multitask architecture grafted upon a shared, pretrained transformer backbone. Through early experimentation, we compared XLM-RoBERTa-base (278M parameters, SentencePiece tokenizer) and mBERT (178M parameters, WordPiece tokenizer). The foundation outputs contextualized embedding states \(\mathbf{h}_t \in \mathbb{R}^{768}\).

\textbf{Causal Mask Injection:}
A critical deficiency natively inherent in modern bidirectional transformers (BERT, RoBERTa) is attention leakage. If evaluating position \(t\), the bidirectional attention matrix fundamentally allows the hidden state to ``look ahead'' to representations at \(t+1\), transforming the strict anticipation task into a trivial detection task. We aggressively countered this by manipulating the \verb|is_decoder=True| parameter and explicitly superimposing a lower-triangular causal attention mask. This replaces all future \(t+k\) attention coefficients with \(-10{,}000.0\), blinding the network to incoming text streams.

\textbf{Multilayer Perceptron Heads:}
Initial iterations utilized simple linear projections (\(\mathbb{R}^{768} \rightarrow \mathbb{R}^{N}\)), which resulted in severe bottlenecks given the 10-language cardinality of the training regime. We substituted these paths with robust 2-layer Multilayer Perceptrons (MLPs). The Switch and Duration heads independently pass the \(h_t\) vector through a dense layer compressing to \(\mathbb{R}^{256}\), followed by a non-linear GELU activation mapping and a Dropout gate (\(p = 0.1\)) before final scalar projection. 

\subsection{Multitask Focal Loss and Burstiness Regularization}
The extreme 4:1 class imbalance required the integration of Focal Loss for the binary switch head logic, which significantly penalized the model from constantly predicting the ``no-switch'' majority class:
\begin{equation}
    \text{FL}(p_t) = -\alpha_t \cdot (1 - p_t)^\gamma \cdot \log(p_t)
\end{equation}
Through hyperparameter tuning ablations, we set \(\alpha = 0.8\) and \(\gamma = 2.0\).

The total loss sums the weighted inputs:
\begin{equation}
    \mathcal{L}_{total} = \lambda_{sw} \cdot \text{FL} + \lambda_{dur} \cdot \text{CrossEntropy}(y^{dur})
\end{equation}

\textbf{Multitask Justification and Burstiness:}
Code-switching is highly ``bursty''; an individual does not merely swap languages dynamically for single random vocabulary words, but rather engages in distinct duration-based cognitive blocks (e.g., lexical insertions vs. deep clausal alternation). By forcing the shared transformer backbone to simultaneously forecast the \textit{duration} of the upcoming switch, we hypothesize an auxiliary regularization effect: the model must construct comprehensive, long-term syntactic representations to differentiate between a brief 1-token insertion and a 7-token sentence finale, rather than simply memorizing localized vocabulary triggers. We validate this hypothesis empirically via a controlled single-task ablation in Section 5.2. We stabilized the network prioritization via \(\lambda_{sw}=0.67\) and \(\lambda_{dur}=0.33\), ensuring the weights sum to unity as a proper convex combination.

\subsection{Training Configuration}
All experiments were conducted using the AdamW optimizer~\cite{conneau2020xlmr} with a linear warmup schedule ramping from 0 to the target learning rate over the first 10\% of training steps, followed by linear decay. The default learning rate was set to \(2 \times 10^{-5}\) with a weight decay of \(0.01\). For the final maximum-supervision runs, we employed early stopping with a patience of 2 epochs monitored on validation loss to prevent overfitting.

Batch sizes ranged from 16 to 64 across ablation experiments, with a default of 32. Gradient accumulation was not required given the moderate sequence lengths (256 subword tokens per sample).

\textbf{Full Fine-Tuning Strategy.}
For our definitive experiments and maximum-supervision ceilings, we employed full fine-tuning of the entire transformer backbone (all 12 encoder layers trainable). This design choice reflects the observation that anticipatory code-switch prediction requires fundamentally different representations from standard masked language modeling: the model must learn to encode forward-looking syntactic boundary signals under causal attention constraints, a capability not present in pretrained checkpoints. Selectively freezing lower layers---a common transfer learning strategy for classification tasks where pretrained features remain useful---risks preserving bidirectional representational biases that conflict with the causal prediction objective. We validate this decision empirically via a controlled freezing ablation in Section 5.3.

Training was executed on NVIDIA GPU hardware via Google Colab (T4) and university HPC clusters. A single full-scale run (10 epochs, 10,000 samples across 10 pairs) required approximately 8--12 hours of GPU compute time. Shorter ablation runs (2--4 epochs, 500--3,000 samples) completed in 10 minutes to 2.5 hours depending on configuration. All experimental outputs---model checkpoints, evaluation JSONs, and visualization plots---were automatically serialized to timestamped directories for full reproducibility.

%%% ---------- 5. EXPERIMENTATION AND RESULTS ----------
\section{Experimentation and Results}

Our empirical methodology systematically escalated from isolated hyperparameter validation up to maximum-supervision limits and definitive zero-shot generalization testing. The experimental pipeline was organized into four sequential groups: Groups A and C served as ablation studies comparing backbone architectures and hyperparameter sensitivity, while Group D executed the definitive maximum-supervision and zero-shot generalization experiments.

\subsection{Backbone Selection: mBERT vs. XLM-RoBERTa}
Before committing computational resources to large-scale training, we conducted systematic backbone comparisons across Groups A (epoch and sample size study) and C (batch size and learning rate study). Both groups trained identical dual-head architectures on both mBERT and XLM-RoBERTa, evaluating Mean Anticipatory F1, Universality $\sigma$, Duration Accuracy, and Switch F1 across all 10 language pairs.

\begin{table}[h]
\centering
\small
\caption{Backbone Comparison: Group A (Epoch \& Sample Size)}
\label{tab:groupA}
\begin{tabular}{llccc}
\toprule
\textbf{Config} & \textbf{Model} & \textbf{F1} & \textbf{$\sigma$} & \textbf{Dur} \\
\midrule
ep=3, s=2K & XLM-R & .668 & .096 & .641 \\
ep=3, s=2K & \textbf{mBERT} & \textbf{.680} & \textbf{.096} & \textbf{.647} \\
\midrule
ep=2, s=500 & XLM-R & .609 & .099 & .608 \\
ep=2, s=500 & \textbf{mBERT} & \textbf{.638} & \textbf{.092} & \textbf{.625} \\
\bottomrule
\end{tabular}
\end{table}

\begin{table}[h]
\centering
\small
\caption{Backbone Comparison: Group C (Batch Size \& LR)}
\label{tab:groupC}
\begin{tabular}{llccc}
\toprule
\textbf{Config} & \textbf{Model} & \textbf{F1} & \textbf{$\sigma$} & \textbf{Dur} \\
\midrule
bs=16, lr=1e-5 & XLM-R & .682 & .103 & .649 \\
bs=16, lr=1e-5 & mBERT & .687 & .100 & .651 \\
\midrule
bs=64, lr=5e-5 & XLM-R & .686 & .103 & .655 \\
bs=64, lr=5e-5 & \textbf{mBERT} & \textbf{.697} & \textbf{.098} & \textbf{.651} \\
\bottomrule
\end{tabular}
\end{table}

Tables~\ref{tab:groupA} and~\ref{tab:groupC} reveal three critical patterns. First, \textbf{mBERT consistently outperformed XLM-RoBERTa in every single configuration}, with margins of 0.01--0.03 in Mean F1 and 0.004--0.02 in $\sigma$. This advantage persisted across all hyperparameter settings, suggesting a fundamental architectural advantage of WordPiece tokenization over SentencePiece for anticipatory code-switch prediction tasks. Second, increasing data volume from 500 to 2,000 samples per pair and epochs from 2 to 3 yielded substantial gains ($+$0.042 F1 for mBERT), confirming that the models were not yet saturated at lower data volumes. Third, the model exhibited \textbf{low sensitivity to batch size and learning rate}: the difference between (bs=16, lr=1e-5) and (bs=64, lr=5e-5) was merely 0.010 F1, indicating robust convergence across a wide hyperparameter range.

Based on this evidence, \textbf{mBERT was selected as the sole backbone for all subsequent Group D experiments}. Additionally, the Group C results guided the final hyperparameter selection: bs=32, lr=2e-5 as the default configuration, balancing computational efficiency with performance.

\subsection{Single-Task vs. Multitask Ablation}
To empirically validate the multitask architecture, we conducted a controlled ablation comparing our full dual-head model against a switch-only single-task variant. The single-task model is architecturally identical but disables the duration head by setting \(\lambda_{dur} = 0\), directing all gradient signal exclusively toward switch prediction. Both models were trained under identical conditions (mBERT, 5 epochs, 3,000 samples per pair, bs=32, lr=2e-5).

\begin{table}[h]
\centering
\small
\caption{Single-Task vs. Multitask Ablation}
\label{tab:ablation}
\begin{tabular}{lcccc}
\toprule
\textbf{Model} & \textbf{F1} & \textbf{Sw F1} & \textbf{$\sigma$} & \textbf{Dur} \\
\midrule
Single (sw only) & \textbf{.681} & \textbf{.567} & .098 & .338 \\
Multi (sw+dur) & .664 & .551 & \textbf{.094} & \textbf{.640} \\
\midrule
\multicolumn{2}{l}{Diff.} & -.016 & -.004 & +.302 \\
\bottomrule
\end{tabular}
\end{table}

\begin{table}[h]
\centering
\small
\caption{Per-Pair F1: Single-Task vs. Multitask}
\label{tab:ablation_perpair}
\begin{tabular}{lccc}
\toprule
\textbf{Pair} & \textbf{Single} & \textbf{Multi} & \textbf{Diff.} \\
\midrule
Chinese--EN & \textbf{.838} & .811 & -.027 \\
Japanese--EN & .815 & \textbf{.818} & +.003 \\
Hindi--EN & \textbf{.749} & .712 & -.037 \\
Korean--EN & \textbf{.741} & .702 & -.039 \\
Russian--EN & \textbf{.690} & .673 & -.017 \\
Arabic--EN & \textbf{.674} & .667 & -.007 \\
German--EN & \textbf{.594} & .556 & -.038 \\
Italian--EN & \textbf{.584} & .576 & -.008 \\
French--EN & .565 & \textbf{.567} & +.002 \\
Spanish--EN & \textbf{.564} & .560 & -.004 \\
\bottomrule
\end{tabular}
\end{table}

The results reveal a nuanced trade-off rather than a clear-cut advantage for either paradigm. The single-task model achieves a modest 2.5\% higher Mean F1 (0.681 vs.\ 0.664), as the full gradient capacity is concentrated exclusively on switch prediction. However, this advantage comes at the complete loss of duration prediction capability---the single-task model's duration accuracy of 0.338 is effectively random chance for a 3-class problem (33.3\%).

Critically, the multitask model achieves two notable advantages. First, it attains a \textbf{lower universality $\sigma$} (0.094 vs.\ 0.098), indicating more uniform performance across typologically diverse pairs. This supports the regularization hypothesis: the auxiliary duration objective prevents the model from over-specializing on easy different-script pairs at the expense of harder same-script pairs. Second, the per-pair analysis (Table~\ref{tab:ablation_perpair}) shows that for same-script Latin pairs---the most challenging category---the performance gap narrows to near-zero (French: +0.002, Spanish: -0.004), suggesting that the duration head specifically benefits the cases where switch prediction is most ambiguous.

From a deployment perspective, the multitask architecture represents a practical Pareto improvement: a single model simultaneously delivers switch prediction within 2.5\% of the single-task ceiling \textit{and} duration estimation at 0.640 accuracy, eliminating the need for a separate duration prediction pipeline.

\subsection{Full Fine-Tuning vs. Selective Freezing}
To validate our full fine-tuning strategy, we compared performance across two backbone adaptation regimes. The selective freezing configuration (top 3 layers unfrozen, bottom 9 frozen) was evaluated under 5 epochs with 3,000 samples per pair, while the full fine-tuning reference (all 12 layers trainable) was obtained from the Group C hyperparameter sweep under 4 epochs with 3,000 samples per pair and a batch size of 64. Both used mBERT with identical focal loss parameters ($\alpha=0.8$, $\gamma=2.0$).

\begin{table}[h]
\centering
\small
\caption{Full Fine-Tuning vs. Selective Freezing (Top-3)}
\label{tab:freeze_ablation}
\begin{tabular}{llcccc}
\toprule
\textbf{Strategy} & \textbf{Config} & \textbf{F1} & \textbf{Sw F1} & \textbf{$\sigma$} & \textbf{Dur} \\
\midrule
Selective (top-3) & 5ep, bs=32 & .664 & .551 & \textbf{.094} & .640 \\
Full FT (all 12) & 4ep, bs=64 & \textbf{.697} & \textbf{.581} & .098 & \textbf{.651} \\
\midrule
\multicolumn{3}{l}{Improvement} & +.030 & +.004 & +.011 \\
\bottomrule
\end{tabular}
\end{table}

Despite having one fewer training epoch, the full fine-tuning configuration yields a substantial $+$3.3\% improvement in Mean Anticipatory F1 (0.697 vs.\ 0.664) and consistent gains across Switch F1 ($+$0.030) and Duration Accuracy ($+$0.011). This gap is particularly notable because the selective freezing run had the advantage of an additional epoch, yet still underperformed. The improvement is observed uniformly across all typological groups, confirming that deep adaptation of the lower transformer layers---which encode fundamental cross-lingual morphological and syntactic representations---is beneficial for the causal prediction objective.

The selective freezing configuration achieves a marginally lower $\sigma$ (0.094 vs.\ 0.098), suggesting that frozen pretrained representations may provide a slight regularization effect on cross-pair variance. However, this comes at a meaningful cost to absolute prediction quality. Given that the primary optimization target is anticipatory F1 rather than variance minimization, full fine-tuning represents the superior strategy and was adopted for all subsequent experiments.

\subsection{The Absolute Performance Ceiling (Supervised Max Training)}
Anticipating precisely when a human cognitive pathway will seamlessly shift between two distinct language structures is statistically chaotic. We aimed to discover the strict mathematical limits of subword-tokenizer architectures when subjected to maximal training volumes.

We expanded the training regimen to 10 distinct language pairs, drawing exactly 10,000 samples per pair (100,000 sentences total). We completely uncapped the transformer restrictions, unfreezing 100\% of the backbone layers. We iterated over 10 epochs using a Patience limit of 2 over the validation loss.

\begin{table}[h]
\centering
\small
\caption{Max-Training Supervised Results (mBERT, 10 ep, 10K/pair).}
\label{tab:supervised}
\begin{tabular}{llcc}
\toprule
\textbf{Pair} & \textbf{Group} & \textbf{F1} & \textbf{Dur} \\
\midrule
Chinese--EN & CJK & 0.869 & 0.672 \\
Japanese--EN & CJK & 0.845 & 0.668 \\
Korean--EN & Hangul & 0.767 & 0.658 \\
\midrule
Hindi--EN & Devanagari & 0.760 & 0.661 \\
Russian--EN & Cyrillic & 0.721 & 0.649 \\
Arabic--EN & Arabic & 0.719 & 0.645 \\
\midrule
German--EN & Latin & 0.610 & 0.641 \\
Italian--EN & Latin & 0.606 & 0.638 \\
French--EN & Latin & 0.590 & 0.635 \\
Spanish--EN & Latin & 0.583 & 0.633 \\
\midrule
\multicolumn{2}{l}{\textbf{Mean F1}} & \textbf{0.707} & \\
\multicolumn{2}{l}{\textbf{Mean Dur Acc}} & & \textbf{0.650} \\
\multicolumn{2}{l}{\textbf{$\sigma$}} & \textbf{0.100} & \\
\bottomrule
\end{tabular}
\end{table}

The resulting hierarchy (Table~\ref{tab:supervised}) starkly illustrates the theoretical limitations of subword multilinguality. Despite 10,000 dense conversational samples directly reinforcing the relationships, Spanish and French F1 metrics hit an absolute ceiling hovering at \(\sim 0.59\). In same-script environments (Latin), the lack of clear orthographic separation actively prevents the token representations from easily differentiating loan words, false cognates, and overlapping morphological prefixes. Conversely, the strict Unicode boundaries intrinsic to CJK mappings drastically ease the detection complexity, catapulting Chinese and Japanese anticipations beyond an F1 of 0.840. The aggregation settled at a mean F1 of 0.707, marking the strict upper performance bound of our dual-head mBERT architecture.

\textbf{Duration Accuracy Analysis.} The duration prediction head achieved a global mean accuracy of 0.650 across all 10 pairs. Notably, duration accuracy exhibits substantially less variance than switch F1 across typological groups: CJK pairs averaged 0.666 while Latin pairs averaged 0.637, a gap of only 0.029 compared to the 0.27 gap in switch F1. This compressed variance suggests that duration prediction operates on partially orthogonal features from switch prediction---once a switch is correctly identified, the temporal extent of the switch segment is governed more by discourse-level patterns (e.g., clause length distributions) than by script-specific cues. The overall plateau at $\sim$0.65 likely reflects the inherent ceiling of the 3-class binning strategy; finer-grained duration labels or regression-based formulations may yield improvements in future work.

\subsection{Zero-Shot Generalization Test}
The true metric defining a viable universal predictor is out-of-distribution inference. If the network merely memorizes the probabilistic collocations of a language (e.g., memorizing that the Spanish article "el" frequently triggers after an English verb), it is not predicting code-switching behavior, but rather memorizing vocabulary statistics.

To aggressively challenge the network's universality, we executed a separate, heavily restricted training run. We fed the model only 6 language pairs (capping volume at 5,000 samples for a brief 5 epochs). Critically, we explicitly omitted Chinese, Japanese, French, and Spanish from the entirety of the training environment. The model was then prompted to perform causal inferences upon these invisible pairings immediately at test time.

\begin{table}[h]
\centering
\caption{Zero-Shot Generalization vs. Supervised Ceilings}
\label{tab:zeroshot}
\begin{tabular}{lccc}
\toprule
\textbf{Test Pair} & \textbf{Max Supervised} & \textbf{Zero-Shot} & \textbf{Diff.} \\
\midrule
French--EN & 0.590 & 0.593 & \textbf{+0.003} \\
Spanish--EN & 0.583 & 0.572 & -0.011 \\
\midrule
Chinese--EN & 0.869 & 0.640 & -0.229 \\
Japanese--EN & 0.845 & 0.621 & -0.224 \\
\bottomrule
\end{tabular}
\end{table}

The resulting evaluation metrics established a phenomenal milestone. Against unseen Latin languages (French and Spanish), the zero-shot predictive F1 scored effectively identical to the absolute maximum mathematical certainty obtained by the supervised instance (registering a negligible variance of \(\sim 0.003\)). 

This parity undeniably confirms that the model conquered Latin-script code-switching anticipation entirely universally. Without experiencing a single token of French training data, the network intuitively parsed cross-linguistic syntactic indicators with an efficacy matching heavy supervision. While CJK evaluations suffered standard performance drops (as expected when substituting distantly typed paradigms mid-flight without phonetic alignments), predicting Chinese/Japanese switches at an F1 of 0.640 completely blind highlights remarkable cross-structural stability. Under zero-shot parameters, the system achieved a highly impressive Universality \(\sigma\) of 0.025 across the unobserved pairings.

\textbf{Why Latin Zero-Shot Matches Supervision.} The near-identical performance on French and Spanish without any training exposure reveals that the model is not memorizing language-specific vocabulary triggers but rather learning generalized syntactic anticipation patterns. Latin-script languages share deep morphological roots and grammatical structures (Romance language family for French/Spanish/Italian). The model, having been trained on German, Italian, and other Latin-script pairs, has already internalized the fundamental challenge of same-script code-switch prediction: detecting subtle syntactic boundary cues in the absence of orthographic signals. French and Spanish, sharing this same challenge profile, are immediately solvable by the same learned representations.

\textbf{Why CJK Zero-Shot Degrades.} The $\sim$0.22 F1 gap for Chinese and Japanese under zero-shot conditions reflects a fundamentally different generalization challenge. CJK scripts employ unique character embeddings that the model has never encountered during training. Unlike Latin pairs where the shared alphabet means shared subword tokens, Chinese Hanzi and Japanese Kanji/Kana occupy entirely disjoint regions of the embedding space. The model must rely exclusively on English-side contextual patterns to infer the switch, without any target-language signal. That it still achieves F1 scores of 0.640 and 0.621 under these extreme conditions demonstrates that English-side syntactic cues alone carry substantial predictive power for code-switching anticipation.

\subsection{Universality Metric Comparison}
We define the Universality Sigma ($\sigma_{universality}$) as the standard deviation of Anticipatory F1 scores across all evaluated language pairs. A lower $\sigma$ indicates more consistent performance, reflecting a model that generalizes uniformly rather than excelling on easy pairs while failing on difficult ones.

\begin{table}[h]
\centering
\caption{Universality $\sigma$ Across Experimental Paradigms}
\label{tab:sigma}
\begin{tabular}{lcc}
\toprule
\textbf{Paradigm} & \textbf{Mean F1} & \textbf{$\sigma_{universality}$} \\
\midrule
Group A (ep=2, s=500, mBERT) & 0.638 & 0.092 \\
Group A (ep=3, s=2000, mBERT) & 0.680 & 0.096 \\
Group C (bs=64, lr=5e-5, mBERT) & 0.697 & 0.098 \\
Max Supervised (10 pairs) & 0.707 & 0.100 \\
\textbf{Zero-Shot (4 held-out)} & \textbf{0.607} & \textbf{0.025} \\
\bottomrule
\end{tabular}
\end{table}

Table~\ref{tab:sigma} reveals a counterintuitive finding: the zero-shot paradigm achieves the \textit{lowest} universality variance despite having the lowest mean F1. Supervised training on more data and more pairs slightly increases $\sigma$ because it amplifies the inherent performance gap between easy (different-script) and hard (same-script) pairs. The zero-shot evaluation, by contrast, was tested only on the 4 held-out pairs which happened to span both script categories, and the model's generalized representations produced remarkably uniform predictions across them. This suggests that the learned anticipatory features are not over-specialized to any particular script type but capture genuinely universal syntactic patterns.

\subsection{Qualitative Error Analysis}
To understand the real-world behavior masked by the rigid F1 aggregation, we conduct a token-level qualitative analysis highlighting realistic streaming interactions across the multi-lingual datasets. 

\textbf{True Positive (Anticipating Clausal Boundaries):}
When dealing with inter-sentential transitions (a complete thought ending before transitioning languages), the model excels natively. 
\begin{itemize}
    \item \textit{Context:} ``The election results were largely unexpected, ''
    \item \textit{Hidden Future Switch:} ``pero la gente estaba feliz.'' (Spanish)
    \item \textit{Prediction:} At the comma token, the model anticipates the termination of the English clause and correctly registers a Switch Probability of \( P = 0.82 \) alongside a Large duration class (\( > 7 \) tokens). The attention matrix effectively keys off grammatical punctuation.
\end{itemize}

\textbf{False Positive (The Loan-Word Cognate Illusion):}
The most prevalent failure mapping in Latin configurations stems from shared linguistic DNA and loan-words that temporarily confuse the subword embeddings.
\begin{itemize}
    \item \textit{Context:} ``I need to buy some ''
    \item \textit{Hidden Future Switch:} ``...jalapeños for the salsa.'' (English continuation)
    \item \textit{Prediction:} The model spikes a Switch Probability to \( P = 0.65 \) directly preceding ``jalapeños'', predicting a small duration burst. While objectively incorrect (the sentence remains grammatically English), the assumption is sociolinguistically highly plausible due to the culturally distinct noun.
\end{itemize}

\textbf{False Negative (Deep Intra-Sentential Swaps):}
When bilingual speakers flip languages seamlessly mid-phrase without terminating the predicate, anticipatory probability crashes heavily.
\begin{itemize}
    \item \textit{Context:} ``We should probably just ''
    \item \textit{Hidden Future Switch:} ``aller au magasin demain.'' (French)
    \item \textit{Prediction:} The model assigns a minimal Switch Probability \( P = 0.12 \) at ``just''. Because English auxiliary verb transitions typically demand an English foundational verb (``go''), the causal backbone lacks sufficient pre-text to justify a drastic language swap without explicit conversational context.

\end{itemize}

\subsection{Limitations}
Several limitations constrain the scope and interpretation of our findings, each suggesting productive directions for future investigation.

\textbf{Same-Script Performance Ceiling.} The most significant architectural limitation is the persistent F1 plateau at $\sim$0.59 for Latin-script language pairs. When both the source and target languages share the same alphabet, subword tokenizers (WordPiece, SentencePiece) frequently produce identical or overlapping token sequences across languages. For example, the English word ``information'' and the French ``information'' map to identical subword decompositions, rendering the model's token-level representations fundamentally ambiguous at language boundaries. This ceiling appears to be an intrinsic property of subword-based architectures rather than a training deficiency, as additional data and epochs provided negligible improvement beyond 3,000 samples per pair.

\textbf{Duration Accuracy Plateau.} The 3-class duration binning (small: 1--2 tokens, medium: 3--6 tokens, large: 7+ tokens) achieves a ceiling of $\sim$0.65 accuracy. This coarse granularity conflates linguistically distinct phenomena---a 7-token phrase switch and a 20-token paragraph switch both receive the same label. Regression-based formulations or finer-grained categorical bins may improve duration modeling, though at the cost of increased class imbalance.

\textbf{Synthetic Data Dependency.} The SwitchLingua dataset, while unprecedented in scale and linguistic diversity, was generated via large language models under explicit linguistic constraints. Although validated by the LinguaMaster quality framework, synthetic code-switching patterns may not fully capture the pragmatic, sociolinguistic, and affective dimensions that drive real-world language alternation. Validation against naturalistic bilingual corpora remains essential.

\textbf{Zero-Shot CJK Gap.} While zero-shot Latin performance matched supervised ceilings, the $\sim$0.22 F1 gap for CJK pairs indicates that truly script-agnostic generalization remains unsolved. The model's reliance on target-language character embeddings for different-script pairs means that zero-shot inference on entirely novel scripts will always suffer from embedding-space blindness.

%%% ---------- 6. CONCLUSION ----------
\section{Conclusion and Future Work}
This study establishes a comprehensive empirical baseline for real-time predictive code-switching anticipation across typologically diverse language pairs. Through systematic experimentation spanning four experimental groups, we mapped the performance landscape of causal multitask transformers on this under-explored task.

Our key findings are threefold. First, mBERT consistently outperformed XLM-RoBERTa across all configurations, establishing it as the preferred backbone for anticipatory code-switch prediction. Second, we identified a strict performance hierarchy dictated by orthographic distance: CJK pairs (F1 $\sim$0.85) $>$ different-script pairs (F1 $\sim$0.73) $>$ same-script Latin pairs (F1 $\sim$0.60), with an overall supervised ceiling of Mean F1 = 0.707. Third, and most critically, our zero-shot evaluation demonstrated near-perfect generalization on unseen Latin pairs (within 0.003 F1 of supervised performance), proving that the learned representations capture genuinely universal syntactic anticipation patterns rather than language-specific vocabulary associations.

The Universality Sigma metric ($\sigma = 0.025$ for zero-shot inference) provides a principled quantification of cross-linguistic consistency, and we propose its adoption as a standard evaluation metric for multilingual code-switching systems.

Several avenues for future work emerge from our findings. Closing the $\sim$0.22 zero-shot CJK gap may require character-aware or byte-level tokenization strategies that bypass the subword embedding bottleneck entirely. Breaking the 0.60 same-script ceiling could benefit from incorporating contextual language identification signals beyond token representations, such as discourse-level features or speaker metadata as explored by Ostapenko et al.~\cite{ostapenko2022speaker}. Transitioning from discriminative to generative architectures---leveraging autoregressive language models with native causal attention---may provide fundamentally richer anticipatory representations. Finally, evaluation on naturalistic (non-synthetic) bilingual corpora, particularly spoken dialogue transcripts, would validate the ecological validity of our predictive framework for real-world deployment in bilingual keyboards, simultaneous translation, and dialogue systems.

%%% ---------- REFERENCES ----------
\bibliographystyle{ACM-Reference-Format}

\begin{thebibliography}{10}

\bibitem{khanuja2020gluecos}
S.~Khanuja, S.~Dandapat, A.~Srinivasan, S.~Sitaram, and M.~Choudhury.
\newblock GLUECoS: An Evaluation Benchmark for Code-Switched NLP.
\newblock In \emph{Proceedings of the 58th Annual Meeting of the Association for Computational Linguistics}, 2020, pp. 3575--3585.

\bibitem{aguilar2020lince}
G.~Aguilar, S.~Kar, F.~Solorio, S.~B. Kumar, T.~M. A. Hassan, and B.~D. Z. V. 
\newblock LinCE: A Centralized Benchmark for Linguistic Code-switching Evaluation.
\newblock In \emph{Proceedings of the 12th Language Resources and Evaluation Conference}, 2020, pp. 1803--1813.

\bibitem{zhang2018fast}
Y.~Zhang, G.~S. Yin, and J.~W. Zhao.
\newblock A Fast, Compact, Accurate Model for Language Identification of Codemixed Text.
\newblock In \emph{Proceedings of the 2018 Conference on Empirical Methods in Natural Language Processing}, 2018, pp. 328--337.

\bibitem{winata2023decades}
G.~I. Winata, S.~Cahyawijaya, F.~Lix, and P.~Fung.
\newblock The Decades Progress on Code-Switching Research in NLP: A Systematic Survey on Trends and Challenges.
\newblock In \emph{Findings of the Association for Computational Linguistics}, 2023, pp. 2936--2978.

\bibitem{mager2019subword}
M.~Mager, Q.~A.~V. Do, and J.~E. N. 
\newblock Subword-Level Language Identification for Intra-Word Code-Switching.
\newblock In \emph{Proceedings of the 2019 Conference of the North American Chapter of the Association for Computational Linguistics}, 2019, pp. 2005--2011.

\bibitem{conneau2020xlmr}
A.~Conneau, K.~Khandelwal, N.~Goyal, V.~Chaudhary, G.~Wenzek, F.~Guzmán, E.~Grave, M.~Ott, L.~Zettlemoyer, and V.~Stoyanov.
\newblock Unsupervised Cross-lingual Representation Learning at Scale.
\newblock In \emph{Proceedings of the 58th Annual Meeting of the Association for Computational Linguistics}, 2020, pp. 8440--8451.

\bibitem{xie2025switchlingua}
P.~Xie, S.~Wang, and R.~G. H.
\newblock SwitchLingua: The First Large-Scale Multilingual and Multi-Ethnic Code-Switching Dataset.
\newblock In \emph{NeurIPS Datasets and Benchmarks Track}, 2025.

\bibitem{samih2016multilingual}
Y.~Samih, T.~A. R. A., and E.~M. N.
\newblock Multilingual Code-switching Identification via LSTM Recurrent Neural Networks.
\newblock In \emph{Proceedings of the Second Workshop on Computational Approaches to Code Switching}, 2016, pp. 50--59.

\bibitem{ostapenko2022speaker}
A.~Ostapenko, S.~Wintner, M.~Fricke, and Y.~Tsvetkov.
\newblock Speaker Information Can Guide Models to Better Inductive Biases: A Case Study On Predicting Code-Switching.
\newblock In \emph{Proceedings of the 60th Annual Meeting of the Association for Computational Linguistics}, 2022, pp. 3853--3867.

\bibitem{winata2021multilingual}
G.~I.~Winata, S.~Cahyawijaya, Z.~Liu, Z.~Lin, A.~Madotto, and P.~Fung.
\newblock Are Multilingual Models Effective in Code-Switching?
\newblock In \emph{Proceedings of the Fifth Workshop on Computational Approaches to Linguistic Code-Switching}, 2021, pp. 142--153.

\end{thebibliography}

\end{document}
